A general and efficient representation of Ancestral Recombination Graphs

### 2023-09-22 paper.tex
Change I -> D

The word ``ARG'' is now widely used to refer to genetic genealogies in
general~\citep[MORE][]{mathieson2020ancestry}, including
varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
This broad usage means there is disagreement on the
exact object being defined. This causes problems, for example,
when evaluating inferred ARGs, and even when defining
what the criteria for success should be.
It also means there is no shared format for interchanging ARG data between
programs, resulting in the assorted negative consequences with which
geneticists are only too familiar~\citep{excoffier2006computer}.
Perhaps most importantly, the direct linkage to
the original stochastic process definition leads to an
overly narrow interpretation of ARGs, which is significantly
out of step with modern methods, and
can lead to missed opportunities in their application and usage.
Such problems go much deeper than pedantic discussion over terminology;
they directly inform what we should be attempting to estimate, how we
assess success, and the expected limitations on the precision
and accuracy of our inferences.

@@

\section*{Genome ARGs}
We define a genome as the complete set of genetic material that a child
inherits from one parent (i.e.\ haploids have one genome, diploids two, etc.).
In this usage a ``genome'' is equivalent to the definition of a
``haplotype'' used by~\cite{shipilina2023origin}, which we avoid because
it already has many interpretations, and in particular carries the implication of
specific mutational patterns [CITES] that persist over
many generations.
We are not concerned here with mutational processes or observed sequences,
% FIXME topological isn't the right word...
but consider only processes of inheritance,
following the standard practice in coalescent theory.
We will also use the term ``genome'' in its
more common sense of ``the genome'' of a species, denoting the full set
of chromosomes and their coordinate spaces, and hope that the distinction
will be clear from the context.

@@

\begin{figure}
\begin{center}
    % FIXME making this 0.75 textwidth for now because it's too tall.
    % https://github.com/tskit-dev/what-is-an-arg-paper/issues/321
    \includegraphics[width=0.75\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
An example genome ARG embedded in a pedigree.
(A) Diploid individuals, visualised in a highly inbred pedigree and
labelled $I_1$ to $I_8$, contain both paternal and maternal  genomes
labelled \textsf{a} to \textsf{p}. Black lines show inheritance paths connecting
genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
Genomes \textsf{a} and \textsf{c} are the product of two independent
recombination events between
the paternal genomes \textsf{e}
and \textsf{f}, and regions of genome inherited are shown.
Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker.
(B) The corresponding gARG along with inheritance annotations on all edges
(partial inheritance in bold).
(C) The corresponding local trees.
}
\end{figure}

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
In this diploid example, each genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \noderef{a} and \noderef{b},
inherited from parents $I_3$ and $I_4$.
Genome \noderef{b} was inherited directly from $I_4$'s genome \noderef{g} without
recombination, whereas
\noderef{a} is the product of
recombining $I_3$'s two genomes \noderef{e} and \noderef{f} at position 2.
This recombination at means \noderef{a} inherits the (half-closed) genome
interval $[0, 2)$ from \noderef{e} and it inherits $[2, 10)$ from \noderef{f};
these intervals are shown attached to the corresponding graph edges.

@@

\section*{Prospective vs retrospective ARGs}
Event ARGs, as defined in the previous section as a way of encompassing
the range of definitions provided in the literature, are
inherently \emph{retrospective}.
Event ARGs are defined in terms
of the events that occur in the history of some set of sampled
sequences.
For example, Fig.~\ref{fig-arg-in-pedigree}
is retrospective because it only includes information
about genetic inheritance relative to the sample individuals
$I_1$ and $I_2$ (genomes \textsf{a}--\textsf{d}).
Genetic transmission must have occurred between (e.g.)
$I_8$ and her daughter $I_6$, but this is not recorded because
% this is probably not true, given we know the complete pedigree?
it is not relevant to the genetic ancestry of the samples.

@@

[NOTE This is a pedantic and slightly bullshitty to end up on,
just writing it down to see if it works. Should probably cut.]
A final point to note, is that
if we are using the term ARG
to loosely refer to genetic genealogies of varying
levels of detail, either inferred or simulated,
it makes little sense to refer ``the'' Ancestral Recombination
Graph (i.e, in  the singular, like ``the tree'' or
``the phylogeny'')
% TODO add more if we keep this
as favoured by some~\citep[e.g.][]{
arenas2013importance,palamara2016argon,hubisz2020inference}.
While there is of course a single ``true'' ARG that describes
the actual history of a sample, 
under parsimony conditions or when sampling from a model there 
will be astronomically
many ARGs that are equally likely given the data.
There are also many different levels of granularity at which we
can reason about recombination, 
and consequently many different types of ARG.
We should embrace this diversity.

>>>>

The word ``ARG'' is now widely used to refer to genetic genealogies in
general~\citep[MORE][]{mathieson2020ancestry}, including
varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
This broad usage means there is disagreement on the
exact object being defined. This causes problems, for example,
when evaluating inferred ARGs, and even when defining
what the criteria for success should be.
It also means there is no shared format for interchanging ARG data between
programs, resulting in the assorted negative consequences with which
geneticists are only too familiar~\citep{excoffier2006computer}.
Perhaps most importantly, the direct linkage to
the original stochastic process definition leads to an
overly narrow interpretation of ARGs, which is significantly
out of step with modern methods and large-scale datasets, and
can lead to missed opportunities in their application and usage.
Such problems go much deeper than pedantic discussion over terminology;
they directly inform what we should be attempting to estimate, how we
assess success, and the expected limitations on the precision
and accuracy of our inferences.

@@

\section*{Genome ARGs}
We define a genome as the complete set of genetic material that a child
inherits from one parent (i.e.\ haploids have one genome, diploids two, etc.).
% Talking about "haplotypes" probably just confuses more
% In this usage a ``genome'' is equivalent to the definition of a
% ``haplotype'' used by~\cite{shipilina2023origin}, which we avoid because
% it already has many interpretations, and in particular carries the implication of
% specific mutational patterns [CITES] that persist over
% many generations.
We are not concerned here with mutational processes or observed sequences,
but consider only processes of inheritance,
following the standard practice in coalescent theory.
We will also use the term ``genome'' in its
more common sense of ``the genome'' of a species, denoting the full set
of chromosomes and their coordinate spaces, and hope that the distinction
will be clear from the context.

@@

\begin{figure}
\begin{center}
    % FIXME making this 0.75 textwidth for now because it's too tall.
    % https://github.com/tskit-dev/what-is-an-arg-paper/issues/321
    \includegraphics[width=0.75\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
[FIXME Update fig to make wider and less tall, and change $I$ to $D$.
Also possibly add some tabular encoding of gARG data.]
An example genome ARG embedded in a pedigree.
(A) Diploid individuals, visualised in a highly inbred pedigree and
labelled $D_1$ to $D_8$,
contain both paternal and maternal  genomes
labelled \textsf{a} to \textsf{p}. Black lines show inheritance paths connecting
genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
Genomes \textsf{a} and \textsf{c} are the product of two independent
recombination events between
the paternal genomes \textsf{e}
and \textsf{f}, and regions of genome inherited are shown.
Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker.
(B) The corresponding gARG along with inheritance annotations on all edges
(partial inheritance in bold).
(C) The corresponding local trees.
}
\end{figure}

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
In this diploid example, each genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, diploid individual $D_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \noderef{a} and \noderef{b},
inherited from parents $D_3$ and $D_4$.
Genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
recombination, whereas
\noderef{a} is the product of
recombining $D_3$'s two genomes \noderef{e} and \noderef{f} at position 2.
This recombination at means \noderef{a} inherits the (half-closed) genome
interval $[0, 2)$ from \noderef{e} and it inherits $[2, 10)$ from \noderef{f};
these intervals are shown attached to the corresponding graph edges.

@@

\section*{Prospective vs retrospective ARGs}
Event ARGs, as defined in the previous section as a way of encompassing
the range of definitions provided in the literature, are
inherently \emph{retrospective}.
Event ARGs are defined in terms
of the events that occur in the history of some set of sampled
sequences.
For example, Fig.~\ref{fig-arg-in-pedigree}
is retrospective because it only includes information
about genetic inheritance relative to the sample individuals
$D_1$ and $D_2$ (genomes \textsf{a}--\textsf{d}).
Genetic transmission must have occurred between (e.g.)
$D_8$ and her daughter $D_6$, but this is not recorded because
% this is probably not true, given we know the complete pedigree?
it is not relevant to the genetic ancestry of the samples.

@@

[NOTE This is a pedantic and slightly bullshitty to end up on,
just writing it down to see if it works. Should probably cut.]
A final point to note, is that
if we are using the term ARG
to loosely refer to genetic genealogies of varying
levels of detail, either inferred or simulated,
it makes little sense to refer ``the'' Ancestral Recombination
Graph (i.e, in  the singular, like ``the tree'' or
``the phylogeny'')
% TODO add more if we keep this. Anyone who says "the ARG is"
as favoured by some~\citep[e.g.][]{
arenas2013importance,palamara2016argon,hubisz2020inference}.
While there is of course a single ``true'' ARG that describes
the actual history of a sample,
under parsimony conditions or when sampling from a model there
will be astronomically
many ARGs that are equally likely given the data.
There are also many different levels of granularity at which we
can reason about recombination,
and consequently many different types of ARG.
We should embrace this diversity.
